\Large \textbf{ Phenotype and gene ontology enrichment as guides for disease modeling in \cel{} }

### 2017-04-17 pea_paper_docs/tex/pea.tex
added corrections suggested by raymond

\section*{Abstract}
\bf{}
Genome-wide experiments have the capacity to generate massive amounts of unbiased
data about an organism. In order to interpret this data, dimensionality reduction
techniques are required. One approach is to annotate genes using controlled languages
and to test experimental datasets for term enrichment using probabilistic methods.
Although gene, phenotype and anatomy ontologies exist for \cel{}, no unified
software offers enrichment analyses of all the ontologies using the same methodology.
Here, we present the WormBase Enrichment Suite, which offers users the
ability to test all nematode ontologies simultaneously. We show that
the WormBase Enrichment Suite provides valuable insight into different
biological problems. Briefly, we show that phenotype enrichment analysis (PEA) can
help researchers identify disease phenologs, phenotypes that are homologous across species,
which can inform disease modeling in \cel{}. The WormBase Enrichment Suite
analysis can also shed light on RNA-seq datasets by showing what
molecular functions are enriched, which phenotypes these functions are implicated
in and what tissues are overrepresented in the dataset. Finally, we explore the
phenotype-anatomy relationship, showing that a small subset of highly specific
tissues are disproportionately likely to cause an Egl phenotype, but inferring
tissue expression from an Egl phenotype is limited to the largest tissues.
]

@@

% A brief tutorial on how to use the command line tool can be found in the SI.
% A brief tutorial on how to interact with the suite using python is available
% as a Jupyter notebook~\cite{Perez2007} in the SI.

>>>>

\section*{Abstract}
\bf{}
Genome-wide experiments have the capacity to generate massive amounts of unbiased
data about an organism. In order to interpret this data, dimensionality reduction
techniques are required. One approach is to annotate genes using unified,
unambiguous vocabularies and to test experimental datasets for term enrichment
using probabilistic methods.
Although gene, phenotype and anatomy ontologies exist for \cel{}, no one
software offers enrichment analyses of all three ontologies.
Here, we present the WormBase Enrichment Suite, which offers users the
ability to test all nematode ontologies simultaneously. We show that
the WormBase Enrichment Suite provides valuable insight into different
biological problems, and that phenotype enrichment analysis (PEA) can
help researchers identify disease phenologs, phenotypes that are homologous across
species, which can inform disease modeling in \cel{}. The WormBase Enrichment Suite
analysis can also shed light on RNA-seq datasets by showing what
molecular functions are enriched, which phenotypes these functions are implicated
in and what tissues are overrepresented in the dataset. Finally, we explore the
phenotype-anatomy relationship, showing that a small subset of highly specific
tissues are disproportionately likely to cause an Egl phenotype, but inferring
tissue expression from an Egl phenotype is limited to the largest tissues.
]

@@

Enrichment analyses are performed as described in~\cite{Angeles-Albores2016}.
Briefly, a static dictionary is constructed for each ontology using the
most recent WormBase release annotation information. This dictionary is generated
using a trimming algorithm described in~\cite{Angeles-Albores2016} to remove
terms that are too general and uninformative as well as terms that are very
specific but do not have many annotations. When the user provides a set of genes
to test against, we use a hypergeometric model to test whether a given term within
an ontology is occurring at a greater frequency than expected by random chance.
